Adaptive group-regularized logistic elastic net regression Supplementary material

### 2019-03-27 docs/manuscript.Rnw
incorporated suggestions by Mark

	Two such studies are introduced in Sections \ref{sec:colorectal} and
	\ref{sec:lymphnode}. The first study is concerned with treatment 
	response prediction for colorectal cancer patients, based on sequenced
	microRNAs. In the second study we attempt to predict lymph node metastases,
	using RNAseq data. In both settings additional information on the 
	features is available in the form of a partitioning. Furthermore, in both
	examples data is sparse, so any extra included information may help to boost
	predictive performance. Inclusion of such feature partitions in 
	the analysis is not straightforward with standard methods.

	At first sight, such an \textit{a priori} grouping of the features suggests 
	the group lasso
	\cite[]{meier_group_2008, yuan_model_2006} or one of its extensions, because
	the group lasso penalises and selects features at the group level. 
	There are two limitations to this approach: the group lasso (a) selects groups
	as a whole instead of single 
	features and (b) does not penalise adaptively: all groups are penalised 
	equally. Extensions such as the sparse group lasso 
	\cite[]{simon_sparse-group_2013} partly deal with (a), but do not address (b).

@@

	Recently, \cite{zhang_novel_2018} introduced VBVS, a method similar to ours in
	spirit. Nevertheless, one clear advantage of our method over VBVS is the
	possibility of adaptive inclusion of external information on the features to
	aid in prediction and feature selection. Bayesian versions of support vector 
	machines have been used in classification problems as well 
	\cite[]{chakraborty_bayesian_2011}, but these methods also lack 
	adaptive inclusion of external information on the features.

	In line with the above, our proposed method is (a) subjective: the 
	hyper-parameters are estimated from the data and not chosen on objective 
	grounds, (b) adaptive: prior information is automatically and subjectively 
	weighed with respect to its informativeness, (c) fast compared to full 
	Bayesian analysis or cross-validation, and (d) requires minimal input by the 
	user: only the data and grouping of the features is required. We emphasize that 
	our method is different from recent developments in grouped penalized 
	regression such as \cite{dondelinger_joint_nodate}, where instead of the 
	features, the observations are partitioned.

	The rest of the paper is structured as follows. We introduce the Bayesian 
	logistic generalised elastic net model in Section
	\ref{sec:model}. In Section \ref{sec:estimation} we derive a variational
	approximation to this model and use this novel approximation in the empirical
	Bayes estimation of multiple, group-specific penalty parameters. In Section
	\ref{sec:simulations} we compare the method in a simulation study. In Sections
	\ref{sec:colorectal} and \ref{sec:lymphnode}, we demonstrate the approach in 
	two data sets. In Section \ref{sec:discussion} 
	we conclude by both discussing some differences and
	parallels between the proposed method, termed \texttt{gren}, and other methods
	in the literature.

@@

	In the Bayesian framework, estimation of penalty parameters may be avoided by 
	the addition of a hyperprior to the model hierarchy. The hyperprior takes the 
	uncertainty in the penalty parameters into account by integrating over them. 
	This approach introduces two issues. Firstly, the choice of hyperprior is not 
	straightforward. Many authors suggest a hyperprior from the gamma family of 
	distributions 
	\cite[]{alhamzawi_bayesian_2017,mallick_bayesian_2013,kyung_penalized_2010}, 
	but the precise parametrisation of this gamma prior is not so obvious. A 
	second issue is the loss of correspondence between the  Bayesian and 
	frequentist elastic net. If the ultimate goal is feature selection, this 
	correspondence may be exploited through the automatic feature selection 
	property of the frequentist elastic net. Endowing the penalty parameters with 
	a hyperprior obstructs their point estimation and, consequently, impedes 
	automatic feature selection. Therefore, to circumvent the problem of 
	hyperprior choice and allow for feature selection by the frequentist elastic 
	net, we propose to estimate the penalty parameters by empirical Bayes.

@@

	\subsection{Empirical-variational Bayes}\label{sec:empvarbayes}
	Variational Bayes was shown to underestimate the posterior variance of the 
	parameters, both numerically and theoretically, in several settings 
	\cite[]{rue_approximate_2009,consonni_mean-field_2007,bishop_pattern_2006,
	wang_inadequacy_2005}. This coincides with our experience that the global 
	penalty parameters $\lambda_1$ and $\lambda_2$ tend to be overestimated. To 
	prevent overestimation we use the parametrisation of the elastic net in 
	\cite{friedman_regularization_2010} as introduced in Section 
	\ref{sec:empiricalbayes}: we fix $\alpha$ and estimate $\lambda$ by 
	cross-validation of the regular elastic net model, such that the overall 
	penalisation is determined by cross-validation of only $\lambda$. By combining
	cross-validation of the global penalty parameter $\lambda$ with empirical 
	Bayes estimation of the penalty multipliers $\blambda' = \begin{bmatrix} 
	\lambda'_1 & \cdots & \lambda'_G \end{bmatrix} \tr$, the estimation is more 
	robust to underestimation of the variational posterior variances. The 
	remaining issue is the choice of $\alpha$. \cite{hastie_glmnet_2016} recommend
	to either fix $\alpha$ \textit{a priori}, or compare the results for several 
	choices of $\alpha$. We recommend the latter.

@@

  We compared the performance of \texttt{gren} to ridge, \texttt{GRridge}, 
  elastic net, \texttt{SGL}, \texttt{cMCP}, and \texttt{gel}. Of the latter 
  three methods, we only present \texttt{cMCP} here; the results for 
  \texttt{SGL} and \texttt{gel} are presented in the SM Section 
  [! section ref here]. For the methods with a
  tuning parameter $\alpha$, we picked the $\alpha$('s) that gave the best
  performance and refer
  the reader to SM Section \ref{sm-sec:colorectal} for comparisons with 
  different $\alpha$'s.

@@

  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}a). 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes in terms of AUC 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}b). 
  This is especially pronounced for the larger models. The performance of
  \texttt{cMCP} is somewhat unstable. This unstable estimation is an issue for   
  all group lasso extensions that we investigated.
  (SM Section \ref{sm-sec:colorectal}). Overall, it seems as though the 
  inclusion of the extra information on the 
  features benefited predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit it only slightly for \texttt{GRridge}.
<<barplot_lines_micrornaseq_colorectal_cancer_res1_auc, fig.cap="Estimated (a) penalty multipliers and (b) AUC in the colorectal cancer example.", out.width="100%", fig.asp=1/2>>=
@

  \section{Application to RNAseq in lymph node cancer}\label{sec:lymphnode}
  The aim of our second example study was to predict lymph node metastasis in
  oral cancer patients using sequenced RNAs from TCGA 
  \cite[]{the_cancer_genome_atlas_network_comprehensive_2015}. The features are 
  3096 transformed and normalised TCGA RNASeqv2 profiles of head-and-neck 
  squamous cell carcinomas on 133 patients with HPV-negative tumours in the oral
  cavity. Of these patients, 76 suffered from lymph node metastasis, while 57 
  did not. For a thorough introduction of these data, see 
  \cite{te_beest_improved_2017}.

  We considered two sources of external feature information: (a) the 
  cis-correlations between the RNASeqv2 data and TCGA DNA copy numbers on 
  the same patients, quantified by Kendall's $\tau$ and binned into five groups.
  In addition, we used (b) $p$-values from a microarray data set
  described in \cite{mes_prognostic_2017}, again binned into five groups. 
  We expected features 
  with a large positive Kendall's $\tau$ to be more important in 
  metastasis prediction \cite[]{masayesva_gene_2004}. Likewise, we expect 
  features with low external $p$-values to be more important.

  We compared \texttt{gren} to the same methods as in Section 
  \ref{sec:colorectal}. However, since the feature information
  consists of two partitions with overlapping groups, we used an extension of
  \texttt{cMCP} that allows for
  overlapping groups \cite[]{zeng_overlapping_2016}.
  We estimated AUC for the methods on an independent validation set of 97 
  patients \cite[]{mes_prognostic_2017}. The features 
  in this validation set were
  microarray expression levels, normalised to account
  for a difference in scale compared to the RNAseq training data. We present the
  results on this validation set, together with the estimated penalty 
  multipliers for \texttt{gren} and \texttt{GRridge}, in Figure 
  \ref{fig:barplot_lines_rnaseq_lymph_node_metastasis_res1_auc}. Additional
  comparisons and performance measures are presented in the SM Section 
  \ref{sm-sec:lymphnode}.

@@

@@

	Although our method is essentially a reweighted elastic net and can be
	considered weakly adaptive, it is different from 
	the adaptive lasso \cite[]{van_de_geer_adaptive_2011,huang_adaptive_2008,
	zhang_adaptive_2007,zou_adaptive_2006} and adaptive elastic net 
	\cite[]{zou_adaptive_2009} in the sense that it adapts to external information
	rather than to the primary data. It also differs in the scale of adaptation: 
	in the adaptive lasso and elastic net the adaptive weights are feature 
	specific, while in our case they are estimated on the group level, rendering
	the adaptation more robust. As can be seen from both the simulations in 
	Section \ref{sec:simulations} and the applications to different outcomes 
	(prognostic or diagnostic) and marker types 
	(RNAseq, microRNAs, or metabolomics) in Sections \ref{sec:colorectal}--
	\ref{sec:lymphnode}, and SM Sections \ref{sm-sec:colorectal}--
	\ref{sm-sec:cervical}, adaptation to 
	external data may be beneficial for prediction and feature selection. 
	We believe that this is due to the `borrowing of information' effect: 
	estimates of feature effects that behave similarly are shrunken similarly, 
	yielding overall, better estimates.

@@

	A possible weak point of the proposed method is the double EM loop. The double
	loop increases the chance of ending up in a local optimum. In the applications
	discussed in Sections \ref{sec:colorectal}--
	\ref{sec:lymphnode}, and SM Sections \ref{sm-sec:colorectal}--
	\ref{sm-sec:cervical}, we investigated the occurrence of multiple optima, 
	but never encountered them. This does not guarantee that local optima do not
	occur, but 
	it provides some evidence that local optima are not ubiquitous. In any case,
	bad local optima can often be avoided by an informed choice of starting 
	values. In our 
	experience, reasonable starting values are obtained by running a 
	group-regularized ridge regression and using the estimated penalty multipliers
	as starting values for \texttt{gren}. Alternatively, a skeptical user may set 
	the starting values for all penalty multipliers to one, thereby starting from
	the regular elastic net and adaptively adjusting the penalty parameters to a
	group-regularized model. The latter is the default approach in our \texttt{R}
	package

<<>>=
bench1 <- read.table("results/micrornaseq_colorectal_cancer_bench1.csv", 
                     stringsAsFactors=FALSE)
bench2 <- read.table("results/rnaseq_lymph_node_metastasis_bench1.csv", 
                     stringsAsFactors=FALSE)
times1 <- round(bench1$time*10^(-9), 2)
times2 <- round(bench2$time*10^(-9), 2)
session <- gsub("_", ".", devtools::session_info()$platform$system)
@
  The double EM loop runs the danger of excessive computation time. We have   
  compared computation times to the
  other methods introduced in Sections \ref{sec:colorectal}--
  \ref{sec:lymphnode}. All timings were obtained on a Macbook 
  Pro 2016 running \Sexpr{session}. 
  The timings of \texttt{gren} are competitive. On the colorectal cancer example
  we achieved the following times:
  \Sexpr{times1[c(1:3)]} seconds, with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, 
  respectively. \texttt{GRridge} is 
  slightly faster with \Sexpr{times1[4]} seconds, while \texttt{SGL} was slightly
  slower: \Sexpr{times1[c(5:7)]} seconds, with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$.
  \texttt{cMCP} (\Sexpr{times1[c(8:10)]} seconds) and \texttt{gel} 
  (\Sexpr{times1[c(11:13)]} seconds) are faster than \texttt{gren}. 
  The results for the lymph node cancer example are similar. \texttt{gren} 
  achieved 
  \Sexpr{times2[c(1:3)]} seconds, with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, 
  respectively.
  \texttt{GRridge} was slightly faster with \Sexpr{times2[4]} seconds, 
  while \texttt{SGL} was slightly slower: \Sexpr{times2[c(5:7)]} seconds, with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$.
  \texttt{cMCP} (\Sexpr{times2[c(14:16)]} seconds) and \texttt{gel} 
  (\Sexpr{times2[c(17:19)]} seconds) are again faster than \texttt{gren}. 
  We note that a considerable speedup may be achieved by iterating between 
  single EB and single VB updates, instead of running the VB EM 
  until convergence. This works in practice, because an EM tends to take a few 
  initial large steps towards the optimum, followed by a series of smaller 
  steps. We argue that the first few iterations of the VB EM render the 
  approximate expectation in \ref{eq:approximatedloglikelihood}
  accurate enough for a 
  reliable EB step.

@@

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{http://pub.math.leidenuniv.nl/~munchmm/}. All results may be recreated
	from \url{https://github.com/magnusmunch/gren}.

	\section*{Acknowledgements}
	\textit{Conflict of Interest}: None declared. We thank the editors and 
	referees for their useful suggestions and comments.

>>>>

	Two such studies are introduced in Sections \ref{sec:colorectal} and
	\ref{sec:oral}. The first study is concerned with treatment 
	response prediction for colorectal cancer patients, based on sequenced
	microRNAs. In the second study we attempt to predict lymph node metastases,
	using RNAseq data. In both settings additional information on the 
	features is available in the form of a partitioning. Furthermore, in both
	examples data is sparse, so any extra included information may help to boost
	predictive performance. Inclusion of such feature partitions in 
	the analysis is not straightforward with standard methods.

	At first sight, such an \textit{a priori} grouping of the features suggests 
	the group lasso
	\cite[]{meier_group_2008, yuan_model_2006} or one of its extensions, because
	the group lasso penalises and selects features at the group level. 
	There are two limitations to this approach: the group lasso (a) selects entire
	groups instead of single 
	features and (b) does not penalise adaptively: all groups are penalised 
	equally. Extensions such as the sparse group lasso 
	\cite[]{simon_sparse-group_2013} deal with (a), but do not address (b).

@@

	Recently, \cite{zhang_novel_2019} introduced VBVS, a method similar to ours in
	spirit. Nevertheless, one clear advantage of our method over VBVS is the
	possibility of adaptive inclusion of external information on the features to
	aid in prediction and feature selection. Bayesian versions of support vector 
	machines have been used in classification problems as well 
	\cite[]{chakraborty_bayesian_2011}, but these methods also lack 
	adaptive inclusion of external information on the features.

	In line with the above, our proposed method is (a) subjective: the 
	hyper-parameters are estimated from the data and not chosen on objective 
	grounds, (b) adaptive: prior information is automatically and subjectively 
	weighed with respect to its informativeness, (c) fast compared to full 
	Bayesian analysis or cross-validation, and (d) requires minimal input by the 
	user: only the data and grouping of the features is required. We emphasize that 
	our method is different from recent developments in grouped penalized 
	regression such as \cite{dondelinger_joint_2018}, where instead of the 
	features, the observations are partitioned.

	The rest of the paper is structured as follows. We introduce the Bayesian 
	logistic generalised elastic net model in Section
	\ref{sec:model}. In Section \ref{sec:estimation} we derive a variational
	approximation to this model and use this novel approximation in the empirical
	Bayes estimation of multiple, group-specific penalty parameters. In Section
	\ref{sec:simulations} we compare the method in a simulation study. In Sections
	\ref{sec:colorectal} and \ref{sec:oral}, we demonstrate the approach in 
	two data sets. In Section \ref{sec:discussion} 
	we conclude by both discussing some differences and
	parallels between the proposed method, termed \texttt{gren}, and other methods
	in the literature.

@@

	In the Bayesian framework, estimation of penalty parameters may be avoided by 
	the addition of a hyperprior to the model hierarchy. The hyperprior takes the 
	uncertainty in the penalty parameters into account by integrating over them. 
	This approach introduces two issues. Firstly, the choice of hyperprior is not 
	straightforward. Many authors suggest a hyperprior from the gamma family of 
	distributions 
	\cite[]{alhamzawi_bayesian_2018,nengjun_yi_bayesian_2013,
	kyung_penalized_2010}, 
	but the precise parametrisation of this gamma prior is not so obvious. A 
	second issue is the loss of correspondence between the  Bayesian and 
	frequentist elastic net. If the ultimate goal is feature selection, this 
	correspondence may be exploited through the automatic feature selection 
	property of the frequentist elastic net. Endowing the penalty parameters with 
	a hyperprior obstructs their point estimation and, consequently, impedes 
	automatic feature selection. Therefore, to circumvent the problem of 
	hyperprior choice and allow for feature selection by the frequentist elastic 
	net, we propose to estimate the penalty parameters by empirical Bayes.

@@

	\subsection{Empirical-variational Bayes}\label{sec:empvarbayes}
	Variational Bayes was shown to underestimate the posterior variance of the 
	parameters, both numerically and theoretically, in several settings 
	\cite[]{rue_approximate_2009,consonni_mean-field_2007,bishop_pattern_2006,
	wang_inadequacy_2005}. This coincides with our experience that the global 
	penalty parameters $\lambda_1$ and $\lambda_2$ tend to be overestimated. To 
	prevent overestimation we use the parametrisation of the elastic net in 
	\cite{friedman_regularization_2010} as introduced in Section 
	\ref{sec:empiricalbayes}: we fix $\alpha$ and estimate $\lambda$ by 
	cross-validation of the regular elastic net model, such that the overall 
	penalisation is determined by cross-validation of only $\lambda$. By combining
	cross-validation of the global penalty parameter $\lambda$ with empirical 
	Bayes estimation of the penalty multipliers $\blambda' = \begin{bmatrix} 
	\lambda'_1 & \cdots & \lambda'_G \end{bmatrix} \tr$, the estimation is more 
	robust to underestimation of the variational posterior variances. The 
	remaining issue is the choice of $\alpha$. \cite{hastie_glmnet_2016} recommend
	to either fix $\alpha$ \textit{a priori}, or compare the results for several 
	choices of $\alpha$. We use the latter.

@@

  We compared the performance of \texttt{gren} to ridge, \texttt{GRridge}, 
  elastic net, \texttt{SGL}, \texttt{cMCP}, and \texttt{gel}. Of the latter 
  three methods, we only present \texttt{cMCP} here; the results for 
  \texttt{SGL} and \texttt{gel} are presented in the SM Section 
  [! section ref here]. For the methods with a
  tuning parameter $\alpha$, we picked the $\alpha$ 
  (two $\alpha$'s for \texttt{gren}) that gave the best
  performance and refer
  the reader to SM Section \ref{sm-sec:colorectal} for comparisons with 
  different $\alpha$'s.

@@

  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}a). 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes in terms of AUC 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}b). 
  This is especially pronounced for the larger models. The performance of
  \texttt{cMCP} is somewhat unstable. This unstable estimation is an issue for   
  all group lasso extensions that we investigated.
  (SM Section \ref{sm-sec:colorectal}). Overall, the 
  inclusion of the extra information on the 
  features benefited predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit it only slightly for \texttt{GRridge}.
<<barplot_lines_micrornaseq_colorectal_cancer_res1_auc, fig.cap="Estimated (a) penalty multipliers and (b) AUC in the colorectal cancer example.", out.width="100%", fig.asp=1/2>>=
@

  \section{Application to RNAseq in oral cancer}\label{sec:oral}
  The aim of our second example study is to predict lymph node metastasis in
  oral cancer patients using sequenced RNAs from TCGA 
  \cite[]{the_cancer_genome_atlas_network_comprehensive_2015}. The features are 
  3096 transformed and normalised TCGA RNASeqv2 expression values of 3096 genes 
  for 133 HPV-negative oral tumours. Of the corresponding patients, 76 suffered 
  from lymph node metastasis, while 57 
  did not. For a thorough introduction of these data, see 
  \cite{te_beest_improved_2017}.

  We considered two sources of external feature information: (a) the 
  cis-correlations between the RNASeqv2 data and TCGA DNA copy numbers on 
  the same patients, quantified by Kendall's $\tau$ and binned into five groups.
  In addition, we used (b) $p$-values from an independent microarray data set
  described in \cite{mes_prognostic_2017}, again binned into five groups. 
  We expected features 
  with a large positive Kendall's $\tau$ to be more important in 
  metastasis prediction \cite[]{masayesva_gene_2004}. Likewise, we expect 
  features with low external $p$-values to be more important.

  We compared \texttt{gren} to the same methods as in Section 
  \ref{sec:colorectal}. However, since the feature information
  consists of two partitions with overlapping groups, we used an extension of
  \texttt{cMCP} that allows for
  overlapping groups \cite[]{zeng_overlapping_2016}.
  We estimated AUC for the methods on an independent validation set of 97 
  patients \cite[]{mes_prognostic_2017}. The features 
  in this validation set were
  microarray expression levels, normalised to account
  for a difference in scale compared to the RNAseq training data. We present the
  results on this validation set, together with the estimated penalty 
  multipliers for \texttt{gren} and \texttt{GRridge}, in Figure 
  \ref{fig:barplot_lines_rnaseq_lymph_node_metastasis_res1_auc}. Additional
  comparisons and performance measures are presented in the SM Section 
  \ref{sm-sec:oral}.

@@

  \section{Additional applications}\label{sec:applications}
  Two more applications are presented in SM Sections \ref{sm-sec:alzheimer} and
  \ref{sm-sec:cervical}. The first one is concerned with diagnosis of 
  Alzheimer's from 230 metabolites' expression levels in 174 subjects. We 
  included several sources of extra information on the metabolites. In this 
  example, \texttt{gren} performs worse than the group lasso methods. This is 
  due to the smaller number of features to learn the penalty parameters from. 
  Additionally, the metabolites are highly collinear, which further impairs 
  estimation. The group lasso extensions depend on just one penalty parameter,
  so suffer less from the small number of features.

  In the second extra example, we attempt to diagnose 56 women with cervical 
  lesions using 2576 sequenced microRNAs with the purpose of 
  selecting relevent markers for cervical cancer screening programmes. We use 
  include one partitioning of the features to enhance predictive performance and 
  feature selection. In this example, \texttt{gren} outperforms the 
  regular elastic net with respect to predictive performance for a range of 
  model sizes.

@@

	Although our method is essentially a reweighted elastic net and can be
	considered weakly adaptive, it is different from 
	the adaptive lasso \cite[]{van_de_geer_adaptive_2011,huang_adaptive_2008,
	zhang_adaptive_2007,zou_adaptive_2006} and adaptive elastic net 
	\cite[]{zou_adaptive_2009} in the sense that it adapts to external information
	rather than to the primary data. It also differs in the scale of adaptation: 
	in the adaptive lasso and elastic net the adaptive weights are feature 
	specific, while in our case they are estimated on the group level, rendering
	the adaptation more robust. As can be seen from both the simulations in 
	Section \ref{sec:simulations} and the applications to different outcomes 
	(prognostic or diagnostic) and marker types 
	(RNAseq, microRNAs, or metabolomics) in Sections \ref{sec:colorectal}--
	\ref{sec:oral}, and SM Sections \ref{sm-sec:colorectal}--
	\ref{sm-sec:cervical}, adaptation to 
	external data may be beneficial for prediction and feature selection. 
	We believe that this is due to the `borrowing of information' effect: 
	estimates of feature effects that behave similarly are shrunken similarly, 
	yielding overall, better estimates.

@@

	A possible weak point of the proposed method is the double EM loop. The double
	loop increases the chance of ending up in a local optimum. In the applications
	discussed in Sections \ref{sec:colorectal}--
	\ref{sec:oral}, and SM Sections \ref{sm-sec:colorectal}--
	\ref{sm-sec:cervical}, we investigated the occurrence of multiple optima, 
	but never encountered them. This does not guarantee that local optima do not
	occur, but 
	it provides some evidence that local optima are not ubiquitous. In any case,
	bad local optima can often be avoided by an informed choice of starting 
	values. In our 
	experience, reasonable starting values are obtained by running a 
	group-regularized ridge regression and using the estimated penalty multipliers
	as starting values for \texttt{gren}. Alternatively, a skeptical user may set 
	the starting values for all penalty multipliers to one, thereby starting from
	the regular elastic net and adaptively adjusting the penalty parameters to a
	group-regularized model. The latter is the default approach in our \texttt{R}
	package

<<>>=
bench1 <- read.table("results/micrornaseq_colorectal_cancer_bench1.csv", 
                     stringsAsFactors=FALSE)

bench2 <- read.table("results/rnaseq_lymph_node_metastasis_bench1.csv", 
                     stringsAsFactors=FALSE)
times1 <- round(bench1$time*10^(-9), 2)
times2 <- round(bench2$time*10^(-9), 2)
session <- gsub("_", ".", devtools::session_info()$platform$system)
@
  The double EM loop runs the danger of excessive computation time. In our 
  implementation we have reduced computational time considerably by applying
  some computational tricks (see SM Section 
  \ref{sm-sec:computationalcomplexity}). In addition, the computationally heavy
  parts were implemented in \texttt{C++} to further reduce the computation time. 
  To assess this reduction, we have compared computation times to the
  other methods introduced in Sections \ref{sec:colorectal}--
  \ref{sec:oral}. All timings were obtained on a Macbook 
  Pro 2016 running \Sexpr{session}. 
  The timings of \texttt{gren} are competitive. On the colorectal cancer example
  we achieved the following times:
  \Sexpr{times1[c(1:3)]} seconds, with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, 
  respectively. \texttt{GRridge} and \texttt{SGL} with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$ are 
  slightly faster with \Sexpr{times1[4]} and \Sexpr{times1[c(5:7)]} seconds, 
  respectively.
  \texttt{cMCP} (\Sexpr{times1[c(8:10)]} seconds) and \texttt{gel} 
  (\Sexpr{times1[c(11:13)]} seconds) are faster than \texttt{gren}. 
  The results for the lymph node cancer example are similar. \texttt{gren} 
  achieved 
  \Sexpr{times2[c(1:3)]} seconds, with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, 
  respectively.
  \texttt{GRridge} was slightly faster with \Sexpr{times2[4]} seconds, 
  while \texttt{SGL} was slightly slower: \Sexpr{times2[c(5:7)]} seconds, with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$.
  \texttt{cMCP} (\Sexpr{times2[c(14:16)]} seconds) and \texttt{gel} 
  (\Sexpr{times2[c(17:19)]} seconds) are again faster than \texttt{gren}. 
  % We note that a considerable speedup may be achieved by iterating between 
  % single EB and single VB updates, instead of running the VB EM 
  % until convergence. This works in practice, because an EM tends to take a few 
  % initial large steps towards the optimum, followed by a series of smaller 
  % steps. We argue that the first few iterations of the VB EM render the 
  % approximate expectation in \ref{eq:approximatedloglikelihood}
  % accurate enough for a 
  % reliable EB step.

@@

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{http://pub.math.leidenuniv.nl/~munchmm/}. The results and documents may 
	be recreated from \url{https://github.com/magnusmunch/gren}.

	\section*{Acknowledgements}
	We thank the editor and referees for their useful suggestions and comments. 
	\textit{Conflict of Interest}: None declared.

### 2019-03-27 docs/supplement.Rnw
incorporated suggestions by Mark

  \section{Introduction}\label{sec:introduction}
  This document contains supplementary material to the paper `Adaptive 
  group-regularized logistic elastic net regression'. Section 
  \ref{sec:enetprior} contains extra details on the elastic net prior 
  distribution. Section \ref{sec:objectivefunction} gives the objective 
  function that is maximised in the variational Bayes procedure introduced in
  Section \ref{md-sec:estimation} of the Main Document (MD). Section
  \ref{sec:polyagamma} gives details on
  the P{\'o}lya-Gamma parametrisation introduced in Section 
  \ref{md-sec:estimation} of the MD. The
  derivations of the variational posterior used in the MD are shown in Section
  \ref{sec:variationaldistributions}. Section
  \ref{sec:computationalcomplexity} shows how we may reduce the computational
  complexity of our variational Bayes procedure. Section 
  \ref{sec:expectedloglikelihood} elaborates on the Variational Bayes expected
  joint log likelihood that is maximised to estimate the penalty multipliers as
  explained in Section \ref{md-sec:estimation} of the MD. The full procedure of
  the
  proposed method is shown in Section \ref{sec:fullestimationprocedure}. 
  Additional results to the applications in the MD Sections 
  \ref{md-sec:colorectal}--\ref{md-sec:lymphnode}. We
  apply our method to a study on metabolomics in Alzheimer's in Section
  \ref{sec:alzheimer} and to study in microRNAs in cervical cancer data 
  in Section \ref{sec:cervical}. Finally, 
  we introduce the ridge and lasso versions of the proposed method in Section
  \ref{sec:evbridgelasso}.

@@

  \section{Application to RNAseq in lymph node cancer}
  \label{sec:lymphnode}
  In addition to the results presented in Section \ref{md-sec:lymphnode} of the 
  MD, we compared \texttt{gren} to the sparse group lasso (\texttt{SGL}) by 
  \cite{simon_sparse-group_2013} and the group exponential lasso (\texttt{gel}) 
  by \cite{breheny_group_2015}. The original group lasso methods are not 
  designed to deal with overlapping groups, so in addition to using the versions
  that allow for overlapping groups, we 
  cross-tabulated the two co-data partitions to create one paritioning with 25 
  groups for \texttt{cMCP} and \texttt{gel}.
  We use $\alpha \in \{0.05, 0.5, 0.95 \}$ for the
  methods that require this extra tuning parameter. The AUCs and Brier skill 
  scores calculated on the
  validation set are presented in Figures 
  \ref{fig:lines_rnaseq_lymph_node_metastasis_res1_auc}--
  \ref{fig:lines_rnaseq_lymph_node_metastasis_res1_briers}
<<lines_rnaseq_lymph_node_metastasis_res1_auc, fig.cap="Estimated AUC for the lymph node cancer example.", out.width="100%", fig.asp=1>>=
@

@@

  \section{Application to microRNAs in cervical cancer}\label{sec:cervical}
  A deep sequencing analysis on small non-coding ribonucleic acid (microRNAseq) 
  was performed on 24 women with high-grade cervical intraepithelial 
  neoplasia (CIN3) and 32 healthy women for the purpose of finding relevant 
  screening markers for cervical cancer screening. The next generation 
  sequencing analysis resulted in 2,576 transcripts. The data was normalized 
  and pre-processed, rendering 752 transcripts. More details of the
  data sets and the pre-processing steps are found in the supplementary   
  material of \cite{novianti_better_2017}.

@@

>>>>

  \section{Introduction}\label{sec:introduction}
  This document contains supplementary material to the paper `Adaptive 
  group-regularized logistic elastic net regression'. Section 
  \ref{sec:enetprior} contains extra details on the elastic net prior 
  distribution. Section \ref{sec:objectivefunction} gives the objective 
  function that is maximised in the variational Bayes procedure introduced in
  Section \ref{md-sec:estimation} of the Main Document (MD). Section
  \ref{sec:polyagamma} gives details on
  the P{\'o}lya-Gamma parametrisation introduced in Section 
  \ref{md-sec:estimation} of the MD. The
  derivations of the variational posterior used in the MD are shown in Section
  \ref{sec:variationaldistributions}. Section
  \ref{sec:computationalcomplexity} shows how we may reduce the computational
  complexity of our variational Bayes procedure. Section 
  \ref{sec:expectedloglikelihood} elaborates on the Variational Bayes expected
  joint log likelihood that is maximised to estimate the penalty multipliers as
  explained in Section \ref{md-sec:estimation} of the MD. The full procedure of
  the
  proposed method is shown in Section \ref{sec:fullestimationprocedure}. 
  Additional results to the applications in the MD Sections 
  \ref{md-sec:colorectal}--\ref{md-sec:oral}. We
  apply our method to a study on metabolomics in Alzheimer's in Section
  \ref{sec:alzheimer} and to study in microRNAs in cervical cancer data 
  in Section \ref{sec:cervical}. Finally, 
  we introduce the ridge and lasso versions of the proposed method in Section
  \ref{sec:evbridgelasso}.

@@

  \section{Application to RNAseq in oral cancer}
  \label{sec:oral}
  In addition to the results presented in Section \ref{md-sec:oral} of the 
  MD, we compared \texttt{gren} to the sparse group lasso (\texttt{SGL}) by 
  \cite{simon_sparse-group_2013} and the group exponential lasso (\texttt{gel}) 
  by \cite{breheny_group_2015}. The original group lasso methods are not 
  designed to deal with overlapping groups, so in addition to using the versions
  that allow for overlapping groups, we 
  cross-tabulated the two co-data partitions to create one paritioning with 25 
  groups for \texttt{cMCP} and \texttt{gel}.
  We use $\alpha \in \{0.05, 0.5, 0.95 \}$ for the
  methods that require this extra tuning parameter. The AUCs and Brier skill 
  scores calculated on the
  validation set are presented in Figures 
  \ref{fig:lines_rnaseq_lymph_node_metastasis_res1_auc}--
  \ref{fig:lines_rnaseq_lymph_node_metastasis_res1_briers}
<<lines_rnaseq_lymph_node_metastasis_res1_auc, fig.cap="Estimated AUC for the lymph node cancer example.", out.width="100%", fig.asp=1>>=
@

@@

  \section{Application to microRNAs in cervical cancer}\label{sec:cervical}
  A deep sequencing analysis on small non-coding ribonucleic acid (microRNAseq) 
  was performed on 24 women with high-grade cervical intraepithelial 
  neoplasia (CIN3) and 32 healthy women for the purpose of finding relevant 
  screening markers for cervical cancer screening. The next generation 
  sequencing analysis resulted in 2576 transcripts. The data was normalized 
  and pre-processed, rendering 752 transcripts. More details of the
  data sets and the pre-processing steps are found in the supplementary   
  material of \cite{novianti_better_2017}.

@@

  \section{Gibbs sampler}

  \begin{align*}
    \bbeta | \bomega, \bpsi & \sim \mathcal{N} (\bSigma, \bmu) \\
      & \text{with } \bSigma = [\X \tr \diag (\bomega_i) \X + 
      \lambda_2 \bLambda' + \lambda_2 \bLambda' \diag (\psi_j^{-1})]^{-1} 
      \text{ and } \bmu = \bSigma \X \tr (\y - \frac{1}{2} \mathbf{m}), \\
    \omega_i | \bbeta & \sim \mathcal{PG} (m_i, |\x_i \tr \bbeta|), \\
    \psi_j^{-1} | \beta_j & \sim \mathcal{IG} (\lambda_1/ (2 \lambda_2 
    \sqrt{\lambda'_{g(j)}} |\beta_j|), \lambda_1^2/(4 \lambda_2)).
  \end{align*}
